---
title: "Date Axis Formats"
author: "Dave Lorenz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_height:  1.5
    fig_width: 7
    toc: yes
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Box Plot Examples}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kfigr)
opts_chunk$set(message=FALSE,warning=FALSE,dev="png")
opts_knit$set(eval.after='fig.cap')
```


These examples demonstrate the various date and time axis formats for graphs created using the `timePlot` function. *NOTE:* to use any high-level gaphics function in the smwrGraphs package, you must first call a function to set up the graphics environment like `setPage` or `setPDF`.

The `timePlot` function controls the formatting of the date- (x-) axis with the `xlabels` argument. The default value for `xlabels` is "Auto" but may be any of "hours," "days," "months," "years," or "water years" that will select a specific format. The value for the `xlabels` argument may also be a list of the arguments to the `detePretty` funciton for even more control. The examples in this vignette use primarily the "Auto" option, but at least one example shows one of the other options.

All of these examples focus on the date-axis formatting when the x-axis data is class "Date" or "POSIXt." They plot no data and simplify the y-axis labeling. The x-axis range is set using the `xaxis.range` argument for these graphs, but the range may also be set by the range in the `x` argument to `timePlot` as done in the first example.

```{r }
# Load the smwrGraphs package
library(smwrGraphs)
# Generate the data to plot
DD <- as.POSIXct(c("2009-07-02 10:30", "2009-07-02 12:30"))
YY <- c(0.14, 0.82)
```


#Hour Formats

Hour formats are suitable for date ranges less than about 3 days. In general, hydrologic responses are not substantially shorter time frames than hours, therefore there are no option for axis formats shorter that hours. Minor ticks are by default minutes when the major ticks and labels are hours.

This example demonstrates the hour format for a short time frame, less than one day. It also demonstrates how to add minor ticks to the graph. The `ticks` argument for the hours axis forma can be computed as 60/min - 1, where min is the number of minutes desired between minor ticks.

```{r firstChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('firstChunk'),": Hour format with 15-minute minor ticks.")}
# Set up the graphics environment, the equivalent call for an on screen
#  device could be setPage("square")

AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="")
# add 15-minute minor ticks (3 internal per hour)
addMinorTicks("bottom", AA.pl, 3)
addMinorTicks("top", AA.pl, 3)

```

The next example demnonstrates the hour format suitable for a time frame between about 1 and 3 days. Note that the upper limit for the x-axis range is the day after the day being plotted. That is required because the class "Date" data are always interpreted as midnight of the day. Minor ticks at hourly intervals are added in this example.

```{r secondChunk, anchor="figure", fig.cap=paste0("**Figure ** ", figr('secondChunk'),": The default hour format for a time span of 1 day.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-07-02", "2009-07-03")))
# add hourly minor ticks (5 internal per hour)
addMinorTicks("bottom", AA.pl, 5)
addMinorTicks("top", AA.pl, 5)

```


#Day Formats

Day formats are suitable for date ranges between 1 and several days, generally less than 2 months though. The format of the displayed day changes from the full date to just the day number as the length of the date range increases. Minor ticks are by default hours when the major ticks and labels are days. These examples replicates figure `r figr("secondChunk")`, but forces the format to be days. 

```{r thirdChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('thirdChunk'),": The day format for a time span of 1 day.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-07-02", "2009-07-03")),
                  xlabels="days")
# add hourly minor ticks (5 internal per hour)
addMinorTicks("bottom", AA.pl, 23)
addMinorTicks("top", AA.pl, 23)

```


```{r fourthChunk, anchor="figure", fig.cap=paste0("**Figure ** ", figr('fourthChunk'),": The format for a time span of 1 day with the day label placed between the ticks.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-07-02", "2009-07-03")),
                  xlabels=list(major="days", style="between"))
# add hourly minor ticks
addMinorTicks("bottom", AA.pl)
addMinorTicks("top", AA.pl)

```

The default day labels are placed at the ticks in figure `r figr("thirdChunk")`, but forced to between the ticks in figure `r figr("fourthChunk")`. Hourly ticks are added in both figures; a value for the `ticks` argument is not needed---all minor ticks are drawn.


```{r fifthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('fifthChunk'),": The default format for a time span of 4 days.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-07-01", "2009-07-05")))

```


```{r sixthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('sixthChunk'), ": The default format for a time span of 7 days.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-06-28", "2009-07-05")))

```


```{r seventhChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('seventhChunk'),": The default format for a time span of 28 days.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-06-20", "2009-07-18")))

```

The default x-axis format for time frames longer than 3 days, but shorter than two month is days. Figures `r figr("fifthChunk")`, `r figr("sixthChunk")`, and `r figr("seventhChunk")` show how the format for displaying the day changes as the time frame increases. No ticks are drawn in any of those figures.

#Month Formats

Just as the format of the day changes as the time span chagnes, the format of the month name changes from the full month name to the abbreviated month name to single letters. Single letter abbreviations are not acceptable for U.S. Geological Survey (2012) publication standards, but are included as an option for quick plot review, editing by an illustrator, or others who wish to use them. Minor ticks are by default days when the major ticks and labels are months. To draw minor ticks at specific day intervals requires inlcuding an argument set up like `xlabels=list(major="months", minor="5 days")` in the call to `timePlot`.

```{r eighthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('eighthChunk'),": The default format for a time span of 2 months.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-06-01", "2009-08-01")))
# add daily minor ticks
addMinorTicks("bottom", AA.pl)
addMinorTicks("top", AA.pl)

```


```{r ninthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('ninthChunk'),": The default format for a time span of 1 year.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-06-01", "2010-06-01")))

```


```{r tenthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('tenthChunk'),": The default format for a time span of 18 months.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-01-01", "2010-06-01")))

```


Figures `r figr("eighthChunk")`, `r figr("ninthChunk")`, and `r figr("tenthChunk")` show the change in month format as the time span increases from 2 months to 18 months. Minor ticks indicating days are shown only for figure 8. Note that the format for the month abbreviations in figure 10 can be changed to the standard abbreviation by including `xlabels=list(major="months", label.abbr=2)` in the call to `timePlot`.

#Year Formats

Year formats are suitable for time spans from 1 to many years, although the month format may be used for time spans of a single year. In contrast to the previous formats, the year is always displayed in the 4-digit format, but longer time spans will force the displayed year to be displyed in mutliples of 2, 5, 10, or longer. Minor ticks are by default months when the major ticks and labels are years. 


```{r eleventhChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('eleventhChunk'),": The years format for a time span of 2 years.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2009-01-01", "2011-01-01")),
                  xlabels="year")
# add monthly minor ticks
addMinorTicks("bottom", AA.pl)
addMinorTicks("top", AA.pl)
```


```{r twelthChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('twelthChunk'),": The default format for a time span of 18 months.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2000-01-01", "2013-06-01")))
# add intervening year minor ticks
addMinorTicks("bottom", AA.pl, 1)
addMinorTicks("top", AA.pl, 1)
```


Figures `r figr("eleventhChunk")` and `r figr("twelthChunk")` show the year-axis formats for time spans of 2 years, where the individual years are shown and 12 years which shows the multiple year format. The default format for 2 years is months; figure `r figr("eleventhChunk")` forces the format to be year. Minor ticks are drawn for months in figure `r figr("eleventhChunk")` and the intervening year in figure `r figr("twelthChunk")`.

#Water Year Formats

Thw water year begins on October 1 of the previous year and ends on September 30 of that year. Water Year formats are suitable for time spans from 1 to several years, although the month format may be used for time spans of a single year. Long-term water-year summary data can be displayed using a numeric axis with axis titles, or captions, indicating that the unit of time is the water year. The water year format is never a default, it must always be specified by the `xlabels` argument.

```{r thirteenChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('thirteenChunk'),": The format for a time span of 1 water year.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2008-10-01", "2009-10-01")),
                  xlabels="water year")
# add intervening year minor ticks
addMinorTicks("bottom", AA.pl)
addMinorTicks("top", AA.pl)

```


```{r fourteenChunk, anchor="figure", fig.cap=paste0("**Figure ** ",figr('fourteenChunk'),": The format for a time span of 5 water years.")}
AA.pl <- timePlot(DD, YY, Plot=list(what="none"), ylabels=c(0,1), ytitle="",
                  xaxis.range=as.Date(c("2005-10-01", "2010-10-01")),
                  xlabels="water year")

```


Figures `r figr("thirteenChunk")` and `r figr("fourteenChunk")` show the water year format for 1 and 5 water years. The format changes after a time span of 2 water years. Minor ticks are drawn only in figure 13.

# Bibliography

U.S. Geological Survey, 2012, Author's guide to standards for U.S. 
Geological Survey page-size illustrations, 37 p.


