---
title: "Graph Additions"
author: "Dave Lorenz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_height:  7
    fig_width: 7
    toc: yes
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Graph Additions}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(rmarkdown)
library(captioner)
opts_chunk$set(message=FALSE,warning=FALSE,dev="png")
fig_nums <- captioner()
```


These examples demonstrate how to add features to an existing graph. Adding a plot or a title to an existing graph is discussed in the GraphSetup vignette. These general procedures apply to most high-level graphics functions within the smwrGraphs package. All of the examples use randomly generated sets of data. **NOTE:** to use any high-level gaphics function in the smwrGraphs package, you must first call a function to set up the graphics environment like `setPage` or `setPDF`.

```{r}
# Load the smwrGraphs package
library(smwrGraphs)
# Generate the random data
set.seed(3636)
X <- rnorm(32)
Y <- X + rnorm(32)
```


#Reference Line with Annotation

This example draws a simple scatter plot, then adds a line representing the median y value and annotates that line. Adding annotation generally requires a trail and error approach to placement of the annotation. The simple method used in this example works because X and Y are correlated.

```{r chunk1}
# Set up the graphics environment, the equivalent call for an on screen
#  device would be setPage("square")
setPNG()
AA.pl <- xyPlot(X, Y)
# Add the median line of Y and annotation
refLine(horizontal = median(Y), current=AA.pl)
addAnnotation(min(X), median(Y), "Median Y", current=AA.pl)

```

`r fig_nums('chunk1',"Scatter plot with reference line and annotation.")`

#Grid Lines

Adding grid lines is a 3-step process---1 create the graph with nothing plotted, 2 add the grid lines, and 3 then add the plotted data. This process guarantees that the data will overplot the grid lines.

```{r chunk2}
setPNG()
# Step 1
AA.pl <- xyPlot(X, Y, Plot=list(what="none"))
# Step 2
addGrid(AA.pl)
# Step 3
AA.pl <- addXY(X, Y, Plot=list(what="points"))
```

`r fig_nums('chunk2',"Scatter plot with grid lines.")`

#Add a Smoothed Line

It is sometimes desireable to add a smoothed line to a scatter plot to show the general trend or relation between the data.

```{r chunk3}
setPNG()
# Create a scatter plot from the X and Y data. The name of the output (AA.pl)
#  is completely arbiutrary, but consistently used through these examples.
AA.pl <- xyPlot(X, Y)
# The addSmooth function will compute the smmothed line and add the plot to the
# graph. Accept all defaults for this example. A very useful additional
# argument would be span for the loess.smooth
AA.pl <- addSmooth(X, Y, current=AA.pl)
```

`r fig_nums('chunk3',"Scatter plot with smooth line.")`

#Add a Regression Line with Confidence Intervals and Table

Sometimes a table needs to be added to a graph. That table may represent some
statistical summary or other additional data.

```{r chunk4, fig.height=5}
setPNG()
# Create a scatter plot from the X and Y data. 
AA.pl <- xyPlot(X, Y, Plot=list(what="points", color="black"))
# Create and add the regresion line and 95% confidence intervals
AA.pl <- addSLR(AA.pl)
# The output is discarded in this case becuase an explanation is not created
addCI("SLR", current=AA.pl)
# Create the table and add it to the graph
# Note may actually want to reformat the last p-value so not 0
AA.tbl <- format(round(coefficients(summary(AA.pl$lm)), 4))
AA.tbl <- cbind(" "=c("Intercept", "X"), AA.tbl)
addTable(AA.tbl, where="ul")
```

`r fig_nums('chunk4',"Scatter plot with simple linear regression.")`

#Show area covered by data

Occasionally there is a need to show the general area coverd by a set of data. The `smwrGraphs` package has several functions that help the user draw areas covered by the data: `dataEllipse`, `cov2Ellipse`, and `hull` functions will provide closed polygons that represent some measure of the area covered by data. This example demonstrates `dataEllipse` and `hull`.

This exmaple also identifies and labels selected points lying outside of the enclosing polygon. It demonstrates the use of the `lablePoints` function.

```{r chunk5, fig.height=3.5}
setPNG()
# Create a scatter plot from the X and Y data. 
AA.pl <- xyPlot(X, Y, Plot=list(what="points", color="black"))
# Create and draw an ellipse that covers 90 percent of the data
AA.el <- dataEllipse(X, Y, percent=90)
with(AA.el, addXY(x, y, Plot=list(what="lines", color="darkred"), current=AA.pl))
# Now do the same with a smooth hull
AA.hl <- hull(X, Y, percent=90, smooth=TRUE)
with(AA.hl, addXY(x, y, Plot=list(what="lines", color="magenta"), current=AA.pl))
# Now find the distance from the center of the ellipse and which are greater than 
# the 90th percentile
AA.ds <- mahalanobis(cbind(X,Y), c(mean(X), mean(Y)), var(cbind(X,Y)))
AA.sel <- which(AA.ds > quantile(AA.ds, probs=0.9, type=2))
# Add the labels--the sequence number of the point
labelPoints(X[AA.sel], Y[AA.sel], as.character(AA.sel), current=AA.pl)
```

`r fig_nums('chunk5',"Scatter plot showing area covered by the data.")`

