---
title: "Line/Scatter Examples"
author: "Dave Lorenz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    fig_height:  6
    fig_width: 6
    toc: yes
    fig_caption: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Line/Scatter Examples}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kfigr)
library(rmarkdown)
opts_chunk$set(message=FALSE,warning=FALSE,dev="png")
opts_knit$set(eval.after='fig.cap')
```

These examples demonstrate the basic high-level line/scatter plot functions in the smwrGraphs package. Most of the examples use randomly generated sets of data.  **NOTE:** to use any of these functions, you must first call a function to set up the graphics environment like `setPage` or `setPDF`.

```{r }
# Load the smwrGraphs package
library(smwrGraphs)
# Generate random samples for the examples.
set.seed(2576)
X <- runif(33)
Y <- runif(33)
Z <- rep(c("A", "B", "C"), 11)
X12 <- X[1:12]
Z12 <- LETTERS[1:12]
# Load the smwrData package
library(smwrData)
data(IonBalance)
data(KlamathTP)
```


# Scatter Plot

The `xyPlot` function plots paired x- and y-coordinate data. As of
version 0.7, there are methods for factor and numeric x-coordinate data and numeric y-coordinate data. This example plots only numeric data.

```{r chunk1, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk1'),": A simple scatter plot")}
# Set up the graphics environment, the equivalent call for an on screen
#  device could be setPage("square")

xyPlot(X, Y, Plot=list(color="darkblue"))

```



# Date/Time Plot

The `timePlot` function plots paired x- and y-coordinate data. As of version 0.7, there are methods for Date, POSIXt, numeric, and integer x-coordinate data and numeric y-coordinate data. This example plots total phosphorus concentrations over time.

```{r chunk2, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk2'),": A time-series plot")}
with(KlamathTP, timePlot(sample_dt, TP_ss, Plot=list(what="points", color="darkblue"),
  yaxis.range=c(0,1.5)))
```

# Season Plot

The `seasonPlot` function plots paired x- and y-coordinate data along an x axis that spans one year. The x-coordinate data must be in a form that can be converted to decimal time---representing the date as the year and fractional part of the year. The function `dectime} in `smwrBase} describes that format. This example shows the variation of total phosphorus at a site.

```{r chunk3, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk3'),": A seasonal plot")}
with(KlamathTP, seasonPlot(sample_dt, TP_ss, yaxis.range=c(0,1.5)))

```

# Series Plot

The `seriesPlot` function plots y-coordinate data arranged within a season associated with each observation. The graph shows the sequence of observations within each season over time. The x-axis data is generated from the sequence, so the data to be plotted must be a regular series. The funciton `regularSeries` in `smwrBase` can be used to generate the regular series from data that have irregular sampling. The example plots the total phosphorus data as a regular series. Note that the default method for `seriesPlot` assumes no missing values and if there are any missing values within a season, that season is not plotted.

```{r chunk5, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk5'),": A series plot"), fig.height=4}
# Create the regular series of observations
AA.rs <- with(KlamathTP, regularSeries(TP_ss, sample_dt, begin="1972-01-01", 
  end="1980-01-10"))
seriesPlot(AA.rs$Value, yaxis.range=c(0,1.5), xlabels=month.USGS)
```

# Color Plot

The `colorPlot` function plots paired x- and y-coordinate data, with symbol color controlled by another variable. As of version 0.7, there is only one method, for numeric x and y data. The color variable can be type character or factor, numeric, or integer. If the type is character or factor, then the values can be color names (see the documentation for colors), or group names that are converted to bright colors. If the type is numeric, then the colors are created using the color, groups, and ramp components of the Plot argument. If the type is integer, the the color component of Plot should be "Index" to set the color to the index color of the palette (see the documentation for palette).

```{r chunk6, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk6'),":  A color plot"), fig.height=4}
# Accept the default colors for groups.
AA.pl <- colorPlot(X, Y, color=Z)
addExplanation(AA.pl, where="ul", title="", box.off=FALSE)
```

# Dot Plot

The `dotPlot` function plots continuous x-coordinate data for discrete y-coordinate data. Often, dot plots are used to display single values for each discrete value and often sorted as done in this example.

```{r chunk7, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk7'),":  A dot plot"), fig.height=4, fig.width=4}
# Create the y-axis data as a factor with levels sorted by x
Z12f <- factor(Z12, levels=Z12[order(X12)])
# Plot the results
dotPlot(X12, Z12f)
```



# Scaled Plot

The `scalePlot` function plots x- and y-coordinate data on the same scale. This can be used to assess the relation between x and y in the context of their true magnitudes or for geographic data. This example plots some data in the IonBalance dataset to asses the relative ionic strengths.

```{r chunk8, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk8'),":  A scaled plot")}
# Plot Calcium and Magnesium
with(IonBalance, scalePlot(Ca, Mg, Plot=list(what="points")))
```

# Scatter Plot Matrix

A scatter plot matrix shows multiple pairs of x and y data in separate graphs. This example plots some data in the IonBalance dataset to asses some of the realtions among selected cations. The function `setSplom` must be used to set up the graphics device. For many pairs of data, a very large graphics device can be created. The `Panel` argument is a very flexible way to add statistics or selected lines to each plot.

```{r chunk9, anchor="figure", fig.cap=paste0("**Figure ** ",figr('chunk9'),": A scatter plot matrix."), eval=FALSE}
# Plot Calcium and Magnesium and Sodium
AA.lo <- with(IonBalance, setSplom(num.variables=3, touching=FALSE))
with(IonBalance, splomPlot(cbind(Ca, Mg, Na), Panel=list(line="slr"), layout=AA.lo))

```


